Estimating Carrying Capacity for Chinook Parr using Quantile Random Forest Models

### 2020-06-11 analysis/paper/paper.Rmd
Finished revisions through Methods

<!-- Need to define carrying capacity? -->

@@

Most studies that have investigated fish habitat relationships focus on predicting a species’ distribution (presence / absence) or the average abundance or density, neither of which can be easily manipulated to predict carrying capacity. Further, many of these studies focus on only one or two measures of habitat. @Sweka2010 estimated carrying capacity of Atlantic salmon parr using a quantile regression approach, but the only habitat covariate they included was cumulative drainage area. Traditionally, carrying capacity for salmonids has been estimated through stock-recruitment curves. However, this requires a long time-series of data with variety in the number of spawners which is not usually available [@Cramer2009].

<!-- and it often fails to account for habitat covariates. This manuscript incorporates multiple measures of stream habitat to estimate fish-habitat relationships that encompass the collinear nature of most stream habitat metrics and can be used to predict carrying capacity. Our approach moves across several spatial scales, inferring fish-habitat relationships from detailed, localized habitat data and extrapolating capacity predictions across wide swaths of unsampled locations. -->

In fisheries it has long been recognized that that biotic and abiotic factors limit productivity within and across life-stages. For the purposes of this paper, we define carrying capacity as the maximum number of individuals that can be supported given the quantity and quality of habitat available at a given life-stage. We assume that higher observed relative densities within a given life stage are a function of habitat quantity and quality. Furthermore, we assert that observed fish density is a poor predictor of habitat capacity owing to both a paucity of individuals, especially for threatened or endangered species, and the existence of unmeasured variables that may serve to limit capacity. To address this, we have developed a model to estimate juvenile rearing capacity for Pacific salmon in wadeable streams based on quantile regression forest (QRF) [@Meinshausen2006] models using measurements of fish abundance (and density) and habitat characteristics. QRF models combine the theory and justification of quantile regression modeling [@Koenker1978; @Cade2003] with the flexibility and framework of random forest models [@Breiman2001]. They account for unmeasured variables and can be used to describe the entire distribution of predicted fish densities for a given set of habitat conditions, not just the mean expected density. Random forest models have been shown to outperform more standard parametric models in predicting fish-habitat relationships in other contexts [@Knudby2010]. Quantile regression forests share many of the benefits of random forest models, such as the ability to capture non-linear relationships between independent and dependent variables, naturally incorporate interactions between covariates, and work with untransformed data while being robust to outliers [@Prasad2006]. Meanwhile, quantile regression models have been used in a variety of ecological systems to estimate the effect of limiting factors [@Terrell1996; @Cade2003].

The fish abundance/density and habitat data used to fit the QRF model presented here were available from seven watersheds within the interior Columbia River Basin (CRB), Pacific Northwest, USA. Within the interior CRB two major runs of Chinook salmon (*Oncorhynchus tshawytscha*; hereafter Chinook salmon) occur, stream-type (i.e., spring/summer run) and ocean-type (i.e., fall run), each characterized by different life history characteristics. Stream-type Chinook salmon enter freshwater earlier in the year, spawn in the upper reaches of a watershed, and the juveniles rear for up to 16 months in the freshwater before entering the ocean as smolts. Ocean-type Chinook salmon enter freshwater later (e.g. fall or winter) spawn lower in the watershed, and the juveniles may only spend between several weeks and six months in freshwater before migrating to the ocean. Here we focus on juvenile stream-type Chinook, in particular the summer rearing period during low flow. Data presented here are from Chinook salmon populations in the Upper Columbia River spring-run and Snake River spring/summer-run Evolutionary Significant Units (ESU). The Upper Columbia Spring-run ESU is listed as endangered under the Endangered Species Act, the Snake River Spring/Summer-run is listed as threatened [@NOAA2016; @NOAA2016b].

@@

1. Identify measured habitat characteristics that are most strongly associated with observed Chinook salmon parr abundance and density,
1. Use paired fish and habitat measurements to elicit fish-habitat relationships for those habitat characteristics identified as important for determining fish abundance and density,
1. Predict contemporary habitat carrying capacity at all sites where the important habitat characteristics are measures (i.e., CHaMP sites within the Columbia River Basin),
1. Extrapolate capacity predictions at CHaMP sites across a watershed using globally available attribute data to estimate the Chinook salmon parr capacity of that watershed, and
1. Validate estimates of carrying capacity from QRF across multiple watersheds using independent estimates of capacity (e.g., spawner-recruit relationships).

@@

Fish and habitat data used in our study were collected from eleven watersheds within the interior Columbia River Basin (CRB), Pacific Northwest, USA (Figure \@ref(fig:wtsd-map)). The CRB covers more than 668,000 km$^2$ draining large portions of Idaho, Oregon, and Washington, and smaller portions of Montana, Nevada, Utah, and Wyoming, as well as the southeastern portion of British Columbia. The habitat data used to populate the QRF model were collected by the Columbia Habitat Monitoring Program (CHaMP) [@Volk2017] and were downloaded from https://www.champmonitoring.org. Data from the following `r as.character(english(n_distinct(hab_data$Watershed)))` CHaMP watersheds were used in this study: `r paste(sort(unique(hab_data$Watershed))[-length(unique(hab_data$Watershed))], collapse = ', ')` and `r sort(unique(hab_data$Watershed))[length(unique(hab_data$Watershed))]`. Juvenile density and abundance data were collected in a subset of `r as.character(english(n_distinct(fish_hab$Watershed)))` watersheds, at CHaMP survey reaches and were graciously provided by a number of projects, including the Integrated Status and Effectiveness Monitoring Project [@Volk2017].

@@

CHaMP sites are 200 m to 500 m reaches within wadeable streams across select basins within the interior Columbia River Basin (CRB) selected based on a spatially balanced Generalized Random Tesselation Stratified (GRTS) sample selection algorithm [@Stevens1999; @Stevens2004]. Habitat data within CHaMP sites were collected using the CHaMP protocol [@CHaMP2016] which calls for field data collection during the low-flow period, typically from June through October. CHaMP habitat data include, but are not limited to, measurements describing channel complexity, channel units, disturbance, fish cover, large woody debris, riparian cover, size (depth, width, discharge), substrate, temperature, and water quality.

Juvenile fish surveys were conducted for sp/sum Chinook salmon parr during the summer low-flow season at many of the same sites surveyed using the CHaMP protocol. Survey methods included mark-recapture, three-pass removal sampling, two-pass removal sampling, and single-pass electrofishing, as well as snorkeling. These data were used to estimate sp/sum Chinook salmon parr abundance at all CHaMP sites where fish survey data were available. Three-pass removal estimates used the Carle-Strub estimator [@Carle1978], following advice from @Hedger2013. Two-pass removal estimates used the estimator described by @Seber2002. Mark-recapture estimates used Chapman’s modified Lincoln-Peterson estimator [@Chapman1951] and were deemed valid if they met the criteria described in @Robson1964. These estimates were made using the *removal* function from the *FSA* package [@R-FSA] or the *closedp.bc* function from the *Rcapture* package [@R-Rcapture] in R software [@Rsoftware2019]. Snorkel counts were transformed to abundance estimates using paired snorkel-electrofishing sites to calibrate snorkel counts. For sites with invalid estimates or that were sampled with a single electrofishing pass, we developed an estimate of capture probability based on valid estimates, using a binomial generalized linear mixed effects model. Fixed effects were species, wetted width of the site, density of fish caught on the first pass and all possible two-way interactions. We included a random effect for fish crew / watershed. We used this model to predict abundances based on the number of fish caught on the first pass and any other covariates.

@@

A key step in developing a QRF model to predict fish capacities was selecting the habitat covariates to include in the model. The CHaMP program generated more than 100 habitat metrics at each site, many of which were correlated with each other to one degree or another, as is often the case with stream habitat variables. We sought to include a small set of covariates that were not overly redundant (i.e. not highly correlated), described many aspects of stream habitat (e.g. substrate, temperature, complexity, etc.) and were highly associated with the observed fish densities, presumeably because they contained information about what types of habitat fish sought or avoided. Full details of how covariates were selected can be found in Appendix S1.

@@

Using the selected habitat covariates (Table \@ref(tab:select-mets-table)), we fit a QRF model to predict habitat rearing capacity for spring/summer Chinook salmon parr, during summer months, using the natural log of fish densities as the response. After constructing a random forest, predictions of the mean response can be made by averaging the predictions of all trees, similar to the expected value predictions from a statistical regression model. However, the individual predictions from each tree, viewed collectively, describe the entire distribution of the predicted response, therefore, the random forest model can be used in the same way as other quantile regression methods to predict any quantile of the response. There were missing values for some habitat data; thus, any site visit with more than three missing covariates was removed from the dataset and the remaining missing habitat values were imputed using the *missForest* R package [@R-missForest; @Stekhoven2012]. We fit the QRF models using the *quantregForest* function from the *quantregForest* package [@R-quantregForest] in R software [@Rsoftware2019], incorporating data from `r nrow(qrf_data)` records (paired fish-habitat data) and `r as.character(english(nrow(sel_hab_mets)))` habitat covariates (`r round(nrow(qrf_data) / nrow(sel_hab_mets), 1)` data points per covariate) (Table \@ref(tab:select-mets-table)). The 90th quantile of the predicted distribution was used as a proxy for carrying capacity, following the suggestion of @Sweka2010, and to avoid higher quantiles that draw from the very upper tails of observed fish density, where the variability of predictions may be influenced by small sample size issues.

After model fitting, the QRF model was then be used to predict capacity using measurements of the habitat covariates that were used to fit the model. In our case, this includes all sites within CHaMP basins in the interior Columbia River basin. For CHaMP sites that have been sampled in multiple years, we first calculated the mean for each habitat metric among years to make predictions. In total, we generated `r filter(all_preds, is.na(chnk_per_m_se)) %>% summarise(n_sites = n_distinct(Site)) %>% pull(n_sites)` predictions of spring/summer Chinook salmon parr capacity, during summer months, for the following basins: Entiat, Grande Ronde (including Minam), John Day, Lemhi, Methow, South Fork Salmon, Tucannon, Wenatchee and Yankee Fork.

@@

To predict capacity at larger spatial scales, such as the watershed or population, we developed two extrapolation models (linear and areal capacity) based on globally available attributes (GAA) which were available for the entirety of tributary habitat utilized by a given population. The GAA data used here was taken from the list of GRTS master sample sites that the CHaMP sites were originally selected from [@Larsen2008; @Larsen2016]. Possible covariates included temperature range, growing degree days, an index of disturbance, the square root of cumulative drainage area, stream power, slope, channel type and watershed (Table \@ref(tab:gaa-covars)). The natural log of the CHaMP site capacity predictions was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the *svyglm* function from the *survey* package [@R-survey] in R software [@Rsoftware2019]. We fit two extrapolation models to each type of fish density, one that included watershed as covariate, for use in predicting capacity within CHaMP watersheds, and one that did not, for predicting everywhere else. We then made predictions of linear and areal capacity at all master sample sites throughout the interior CRB, generally spaced about one kilometer apart. Summaries of extrapolation model fit are shown in Table \@ref(tab:extrap-r2-table).

To summarize capacity at larger scales, the mean linear capacity (e.g., fish/m) of the master sample points along a particular tributary is multiplied by the length of that tributary. We first restricted the master sample points and lengths of streams to those with the domain of spring/summer Chinook, as defined by StreamNet ([http://www.streamnet.org](http://www.streamnet.org) ) or using expert opinion from local biologists. The capacities of various tributaries could then be summed to estimate capacity at almost any spatial scale.

@@

Spawner-recruit data from several watersheds within the interior CRB were compiled to validate the extrapolated QRF estimates of spring/summer Chinook salmon parr capacity. Some watersheds had direct estimates of parr, while some had estimates of smolts and pre-smolts from rotary screw traps. For the latter, estimates of parr were calculated using estimates of over-winter survival. A series of spawner-recruit functions were then fit to this data including Beverton-Holt, Ricker, and hockey stick, using the *FSA* package [@R-FSA] in R. Estimates of capacity from each of these spawner-recruit curves were compared with QRF estimates of capacity for the same regions.

@@

We fit a linear regression extrapolation using QRF-based predictions of capacity at all CHaMP sites as the response, and various GAAs as the independent variables. The coefficients for the extrapolation model can be found in Table \@ref(tab:gaa-covars). From this, we have estimates of capacity at every master sample point in the CRB, which are located approximately one kilometer apart.

@@

Given the cost/extent of data necessary for QRF extrapolation in watersheds outside of the CRB, there is a pressing need to develop new tools for habitat analyses.  Unmanned Aerial Systems (UAS or drone, commonly) are gaining popularity in wildlife and ecosystem monitoring for their ease of use, safety, accessibility, and cost-efficiency [@Jones2006; @Chabot2015]. UAS produce high-resolution, permanent data at a fraction of the cost of on-the-ground habitat sampling. Advances in imaging techniques (e.g., multispectral imaging) and post processing (e.g., automation of data collection from imagery) are already demonstrating increase in the efficiency and accuracy of data collection [@Whitehead2014; @LeCun2015; @Weinstein2018].

@@

If a species' carrying capacity is defined or constrained, at least in part, by the habitat in which it lives, then illuminating statistically how such habitat impacts carrying capacity can lead to understanding how a species interacts with its environment. This understanding could be of crucial importance in the realm of conservation, when dealing with an endangered or threatened species, but species/habitat interactions are a core element of ecological studies more generally. We have demonstrated how a quantile regression approach, coupled with a random forest framework, can be used to estimate these relationships, and predict an habitat's capacity. As large ecological datasets become more accessible, and the ability to measure large swaths of habitat more feasible, we believe this approach has many potential applications, from the North American breeding bird survey to groundfish trawl surveys. ***other potential examples (coral reefs? turtles?)***

>>>>

@@

Most studies that have investigated fish-habitat relationships focus on predicting a species’ distribution (presence / absence) or the average abundance or density, neither of which can be easily manipulated to predict carrying capacity. Further, many of these studies focus on only one or two measures of habitat. @Sweka2010 estimated carrying capacity of Atlantic salmon parr using a quantile regression approach, but the only habitat covariate they included was cumulative drainage area. Traditionally, carrying capacity for salmonids has been estimated through stock-recruitment curves. However, this requires a long time-series of data with variety in the number of spawners which is not usually available [@Cramer2009].

<!-- and it often fails to account for habitat covariates. -->

In fisheries it has long been recognized that that biotic and abiotic factors limit productivity within and across life-stages. For the purposes of this paper, we define carrying capacity as the maximum number of individuals that can be supported given the quantity and quality of habitat available at a given life-stage. We assume that higher observed relative densities within a given life stage are a function of habitat quantity and quality. Furthermore, we assert that observed fish density is a poor predictor of habitat capacity owing to both a paucity of individuals, especially for threatened or endangered species, and the existence of unmeasured variables that may serve to limit capacity. To address this, we have developed a model to estimate juvenile rearing capacity for Pacific salmon in wadeable streams based on quantile random forest (QRF) [@Meinshausen2006] models using measurements of fish abundance (and density) and habitat characteristics. QRF models combine the theory and justification of quantile regression modeling [@Koenker1978; @Cade2003] with the flexibility and framework of random forest models [@Breiman2001]. They account for unmeasured variables and can be used to describe the entire distribution of predicted fish densities for a given set of habitat conditions, not just the mean expected density. Random forest models have been shown to outperform more standard parametric models in predicting fish-habitat relationships in other contexts [@Knudby2010]. Quantile regression forests share many of the benefits of random forest models, such as the ability to capture non-linear relationships between independent and dependent variables, naturally incorporate interactions between covariates, and work with untransformed data while being robust to outliers [@Prasad2006]. Meanwhile, quantile regression models have been used in a variety of ecological systems to estimate the effect of limiting factors [@Terrell1996; @Cade2003].

The fish abundance/density and habitat data used to fit the QRF model presented here were available from seven watersheds within the interior Columbia River Basin, Pacific Northwest, USA. Within the interior Columbia River Basin two major runs of Chinook salmon (*Oncorhynchus tshawytscha*) occur, stream-type (i.e., spring/summer run) and ocean-type (i.e., fall run), each characterized by different life history characteristics. Stream-type Chinook salmon enter freshwater earlier in the year, spawn in the upper reaches of a watershed, and the juveniles rear for up to 16 months in the freshwater before entering the ocean as smolts. Ocean-type Chinook salmon enter freshwater later (e.g. fall or winter) spawn lower in the watershed, and the juveniles may only spend between several weeks and six months in freshwater before migrating to the ocean. Here we focus on juvenile stream-type Chinook, in particular the summer rearing period during low flow. Data presented here are from Chinook salmon populations in the Upper Columbia River spring-run and Snake River spring/summer-run Evolutionary Significant Units (ESU). The Upper Columbia spring-run ESU is listed as endangered under the Endangered Species Act, the Snake River spring/summer-run is listed as threatened [@NOAA2016; @NOAA2016b]. Hereafter, we refer to both ESUs simply as Chinook salmon.

@@

1. Identify measured habitat characteristics that are most strongly associated with observed Chinook salmon parr abundance and density,
1. Use paired fish and habitat measurements to elicit fish-habitat relationships for those habitat characteristics identified as important for determining fish abundance and density,
1. Predict contemporary habitat carrying capacity at all sites where the important habitat characteristics are measured (i.e., CHaMP sites within the Columbia River Basin),
1. Extrapolate capacity predictions at CHaMP sites across a watershed using globally available attribute data to estimate the Chinook salmon parr capacity of that watershed, and
1. Validate estimates of carrying capacity from QRF across multiple watersheds using independent estimates of capacity (e.g., spawner-recruit relationships).

@@

Fish and habitat data used in our study were collected from eleven watersheds within the interior Columbia River Basin, Pacific Northwest, USA (Figure \@ref(fig:wtsd-map)). The Columbia River Basin covers more than 668,000 km$^2$ draining large portions of Idaho, Oregon, and Washington, and smaller portions of Montana, Nevada, Utah, and Wyoming, as well as the southeastern portion of British Columbia. The habitat data used to populate the QRF model were collected by the Columbia Habitat Monitoring Program (CHaMP) [@Volk2017] and were downloaded from https://www.champmonitoring.org. Data from the following `r as.character(english(n_distinct(hab_data$Watershed)))` CHaMP watersheds were used in this study: `r paste(sort(unique(hab_data$Watershed))[-length(unique(hab_data$Watershed))], collapse = ', ')` and `r sort(unique(hab_data$Watershed))[length(unique(hab_data$Watershed))]`. Juvenile density and abundance data were collected in a subset of `r as.character(english(n_distinct(fish_hab$Watershed)))` watersheds, at CHaMP survey reaches and were graciously provided by a number of projects, including the Integrated Status and Effectiveness Monitoring Project [@Volk2017].

@@

CHaMP sites are 200 m to 500 m reaches within wadeable streams across select basins within the interior Columbia River Basin selected based on a spatially balanced Generalized Random Tesselation Stratified sample selection algorithm [@Stevens1999; @Stevens2004]. Habitat data within CHaMP sites were collected using the CHaMP protocol [@CHaMP2016] which calls for field data collection during the low-flow period, typically from June through October. CHaMP habitat data include, but are not limited to, measurements describing channel complexity, channel units, disturbance, fish cover, large woody debris, riparian cover, size (depth, width, discharge), substrate, temperature, and water quality.

Juvenile fish surveys were conducted for Chinook salmon parr during the summer low-flow season at many of the same sites surveyed using the CHaMP protocol. Survey methods included mark-recapture, three-pass removal sampling, two-pass removal sampling, and single-pass electrofishing, as well as snorkeling. These data were used to estimate Chinook salmon parr abundance at all CHaMP sites where fish survey data were available. Three-pass removal estimates used the Carle-Strub estimator [@Carle1978], following advice from @Hedger2013. Two-pass removal estimates used the estimator described by @Seber2002. Mark-recapture estimates used Chapman’s modified Lincoln-Peterson estimator [@Chapman1951] and were deemed valid if they met the criteria described in @Robson1964. These estimates were made using the *removal* function from the *FSA* package [@R-FSA] or the *closedp.bc* function from the *Rcapture* package [@R-Rcapture] in R software [@Rsoftware2019]. Snorkel counts were transformed to abundance estimates using paired snorkel-electrofishing sites to calibrate snorkel counts. For sites with invalid estimates or that were sampled with a single electrofishing pass, we developed an estimate of capture probability based on valid estimates, using a binomial generalized linear mixed effects model. Fixed effects were species, wetted width of the site, density of fish caught on the first pass and all possible two-way interactions. We included a random effect for fish crew / watershed. We used this model to predict abundances based on the number of fish caught on the first pass and any other covariates.

@@

A key step in developing a QRF model to predict fish capacities was selecting the habitat covariates to include in the model. The CHaMP program generated more than 100 habitat metrics at each site, many of which were correlated with each other to one degree or another, as is often the case with stream habitat variables. We sought to include a small set of covariates that were not overly redundant (i.e. not highly correlated), described many aspects of stream habitat (e.g. substrate, temperature, complexity, etc.) and were highly associated with the observed fish densities, presumably because they contained information about what types of habitat fish sought or avoided. Full details of how covariates used in the QRF model were selected can be found in Appendix S1.

@@

Using the selected habitat covariates (Table \@ref(tab:select-mets-table)), we fit a QRF model to predict habitat rearing capacity for Chinook salmon parr, during summer months, using the natural log of fish densities as the response. After constructing a random forest, predictions of the mean response can be made by averaging the predictions of all trees, similar to the expected value predictions from a statistical regression model. However, the individual predictions from each tree, viewed collectively, describe the entire distribution of the predicted response, therefore, the random forest model can be used in the same way as other quantile regression methods to predict any quantile of the response. There were missing values for some habitat data; thus, any site visit with more than three missing covariates was removed from the dataset and the remaining missing habitat values were imputed using the *missForest* R package [@R-missForest; @Stekhoven2012]. We fit the QRF models using the *quantregForest* function from the *quantregForest* package [@R-quantregForest] in R software [@Rsoftware2019], incorporating data from `r nrow(qrf_data)` records (paired fish-habitat data) and `r as.character(english(nrow(sel_hab_mets)))` habitat covariates (`r round(nrow(qrf_data) / nrow(sel_hab_mets), 1)` data points per covariate) (Table \@ref(tab:select-mets-table)). The 90th quantile of the predicted distribution was used as a proxy for carrying capacity, following the suggestion of @Sweka2010, and to avoid higher quantiles that draw from the very upper tails of observed fish density, where the variability of predictions may be influenced by small sample size issues.

After model fitting, the QRF model was then be used to predict capacity using measurements of the habitat covariates that were used to fit the model. In our case, this includes all sites within CHaMP basins in the interior Columbia River basin. For CHaMP sites that have been sampled in multiple years, we first calculated the mean for each habitat metric among years to make predictions. In total, we generated `r filter(all_preds, is.na(chnk_per_m_se)) %>% summarise(n_sites = n_distinct(Site)) %>% pull(n_sites)` predictions of Chinook salmon parr capacity, during summer months, for the following basins: Entiat, Grande Ronde (including Minam), John Day, Lemhi, Methow, South Fork Salmon, Tucannon, Wenatchee and Yankee Fork.

@@

To predict capacity at larger spatial scales, such as the watershed or population, we developed two extrapolation models (linear and areal capacity) based on globally available attributes (GAA) which were available for the entirety of tributary habitat utilized by a given population. The GAA data used here was taken from the list of Generalized Random Tesselation Stratified master sample sites that the CHaMP sites were originally selected from [@Larsen2008; @Larsen2016]. Possible covariates included temperature range, growing degree days, an index of disturbance, the square root of cumulative drainage area, stream power, slope, channel type and watershed (Table \@ref(tab:gaa-covars)). The natural log of the CHaMP site capacity predictions was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the *svyglm* function from the *survey* package [@R-survey] in R software [@Rsoftware2019]. We fit two extrapolation models to each type of fish density, one that included watershed as covariate, for use in predicting capacity within CHaMP watersheds, and one that did not, for predicting everywhere else. We then made predictions of linear and areal capacity at all master sample sites throughout the interior Columbia River Basin, generally spaced about one kilometer apart. Summaries of extrapolation model fit are shown in Table \@ref(tab:extrap-r2-table).

To summarize capacity at larger scales, the mean linear capacity (e.g., fish/m) of the master sample points along a particular tributary is multiplied by the length of that tributary. We first restricted the master sample points and lengths of streams to those with the domain of spring/summer Chinook salmon, as defined by StreamNet ([http://www.streamnet.org](http://www.streamnet.org) ) or using expert opinion from local biologists. The capacities of various tributaries could then be summed to estimate capacity at almost any spatial scale.

@@

Spawner-recruit data from several watersheds within the interior Columbia River Basin were compiled to validate the extrapolated QRF estimates of spring/summer Chinook salmon parr capacity. Some watersheds had direct estimates of parr, while some had estimates of smolts and pre-smolts (i.e. fall emigrants) from rotary screw traps. For the latter, estimates of parr were calculated using estimates of over-winter survival. A series of spawner-recruit functions were then fit to this data including Beverton-Holt, Ricker, and hockey stick, using the *FSA* package [@R-FSA] in R. Estimates of capacity from each of these spawner-recruit curves were compared with QRF estimates of capacity for the same regions.

@@

We fit a linear regression extrapolation using QRF-based predictions of capacity at all CHaMP sites as the response, and various GAAs as the independent variables. The coefficients for the extrapolation model can be found in Table \@ref(tab:gaa-covars). From this, we have estimates of capacity at every master sample point in the Columbia River Basin, which are located approximately one kilometer apart.

@@

Given the cost/extent of data necessary for QRF extrapolation in watersheds outside of the Columbia River Basin, there is a pressing need to develop new tools for habitat analyses.  Unmanned Aerial Systems (UAS or drone, commonly) are gaining popularity in wildlife and ecosystem monitoring for their ease of use, safety, accessibility, and cost-efficiency [@Jones2006; @Chabot2015]. UAS produce high-resolution, permanent data at a fraction of the cost of on-the-ground habitat sampling. Advances in imaging techniques (e.g., multispectral imaging) and post processing (e.g., automation of data collection from imagery) are already demonstrating increase in the efficiency and accuracy of data collection [@Whitehead2014; @LeCun2015; @Weinstein2018].

@@

If a species' carrying capacity is defined or constrained, at least in part, by the habitat in which it lives, then illuminating statistically how such habitat impacts carrying capacity can lead to understanding how a species interacts with its environment. This understanding could be of crucial importance in the realm of conservation, when dealing with an endangered or threatened species, but species/habitat interactions are a core element of ecological studies more generally. We have demonstrated how a quantile regression approach, coupled with a random forest framework, can be used to estimate these relationships, and predict an habitat's capacity. As large ecological datasets become more accessible, and the ability to measure large swaths of habitat more feasible, we believe this approach has many potential applications, from the North American breeding bird survey to groundfish trawl surveys. The framework could also be applied to depleted, non-migratory, and isolated populations (e.g. desert pupfish *Cyprinodon macularius*) to identify limiting factors in populations and/or determine whether a given habitat patch could support a viable population if limiting factors were addressed. Capacity estimates could also be used to potential translocation sites to determine if given sites can support an abundance considered viable before investing in translocation efforts.  ***other potential examples (coral reefs? turtles?)***
